### 2019-03-14 2-Chapters/6-Chapter/nonstatbandits.git/NonStatB.tex
Done for now, but I still have to do one plot that Emilie wanted to see.

The Generalized Likelihood Ratio statistic for this test is defined by
\begin{equation}\label{eq:6:firstDefGLRT}
    \GLR(n) :=\frac{\sup\limits_{\mu_0,\mu_1,\tau < n}\ell(X_1, \ldots, X_n ; \mu_0,\mu_1,\tau)}{\sup\limits_{\mu_0}\ell(X_1, \ldots, X_n ; \mu_0)},
\end{equation}
%
where $\ell(X_1, \ldots, X_n ; \mu_0)$ and $\ell(X_1, \ldots, X_n ; \mu_0,\mu_1,\tau)$ respectively denote the likelihoods of the first $n$ observations under a model in $\cH_0$ and $\cH_1$.
High values of this statistic tend to indicate rejection of $\cH_0$ over $\cH_1$.
%
By using the form of the likelihood for Bernoulli distributions, this statistic can be written with the binary relative entropy $\kl$, defined for $x,y\in[0,1]$ (with the usual convention that $t \log(t) = 0$ if $t=0$).
%
\begin{equation}\label{eq:6:BernoulliDivergence}
    \kl(x,y) := x \ln\left(\frac{x}{y}\right) + (1-x)\ln\left(\frac{1-x}{1-y}\right).
\end{equation}
%
Indeed, as proved in Appendix~\ref{app:6:GLR_with_kl} (see also equation $(1.9)$ from \cite{LaiXing10}), we can show that in any one-dimensional exponential family, we have
\[\GLR(n) = \exp\Bigl[ \sup_{s \in [2,n-1]} \left[s \times \kl\left(\widehat{\mu}_{1:s},\widehat{\mu}_{1:n}\right) + (n-s) \times \kl\left(\widehat{\mu}_{s+1:n},\widehat{\mu}_{1:n}\right)\right], \Bigr] \]
%
where for $k \leq k'$, $\widehat{\mu}_{k:k'} := \frac{1}{k - k' + 1} \sum_{l=k}^{k'} X_l$ denotes the average of the observations $X_l$ collected between the instants $k$ and $k'$.
%
This motivates the following definition of the Bernoulli GLRT change point detector.

@@

\TODOL{Include in Appendix a formal statement and a proof of Pinsker's inequality?}

\begin{lemma}\label{lem:6:Delay}
    Let be $\bP_{\mu_0,\mu_1,\tau}$ a probabilistic model under which $X_t \in [0,1]$, and $X_t$ has an average of $\mu_0$ for all $t \leq \tau$, and $\mu_1$ for all $t > \tau$, with $\mu_0 \neq \mu_1$,
    and let $\Delta = |\mu_0 - \mu_1|$.
    Then the B-GLRT test given in Definition~\ref{def:6:GLRDef} satisfies
    \begin{equation}
        \bP_{\mu_0,\mu_1,\tau} (\widehat{\tau}_{\delta} \geq \tau + u) \leq \exp\left( -\frac{2\tau u}{\tau + u}\left(\max\left[ 0, \Delta - \sqrt{\frac{\tau + u}{2\tau u} \beta(\tau + u,\delta)} \right]\right)^2 \right).
    \end{equation}
\end{lemma}

\begin{proof}
    \TODOL{Write it ?}
\end{proof}

@@

Also note that, as any test based on scan-statistics, the B-GLRT can be costly to implement as at every time step, it considers all previous time steps as a possible positions for a change-point. Thus, in practice the following adaptation may be interesting, based on down-sampling the possible time steps:
%
\begin{equation}\label{def:6:GLRTricks}
    \widetilde{\tau}_{\delta} = \inf \left\{ n \in \cN : \sup_{s \in \cS_n} \left[s \times d\left(\widehat{\mu}_{1:s},\widehat{\mu}_{1:n}\right) + (n-s) \times d\left(\widehat{\mu}_{s+1:n},\widehat{\mu}_{1:n}\right)\right] \geq \beta(n,\delta)\right\},
\end{equation}
%
for subsets $\cN$ and $\cS_n$. Following the proof of Lemma~\ref{lem:6:FalseAlarm}, we can easily see that this variant enjoys the same false-alarm control.
However, the detection delay may be slightly increased.
In Appendix~\ref{sub:6:IdeasOptimizations} we show that using these practical speedups has little impact on the regret.

@@

%-----------------------------------------------------------------------------
\paragraph{Details about the algorithm.}

@@

\begin{proof}
    We consider one arm $i\in\{1,\dots,K\}$, and when the \GLRklUCB{} algorithm is running,
    we consider two time steps $s\leq t\in\N^*$, chosen between two restart times for that arm $i$.
    Lines~$3$-$4$ state that $A_u = u \mod \lceil \frac{K}{\alpha} \rceil$
    if $u \mod \lceil \frac{K}{\alpha} \rceil \in \{1,\dots,K\}$
    (see Algorithm~\ref{algo:6:GLRklUCB} for details).
    Thus we have
    \begin{align*}
        n_i(t) - n_i(s)
        & = \sum_{u=s+1}^{t} \indic(A_u = i) \\
        & \geq \sum_{u=s+1}^{t} \indic\left(A_u = i, A_u = u \mod \left\lceil \frac{K}{\alpha} \right\rceil\right) \\
        & \geq \sum_{u=s+1}^{t} \indic\left(i = u \mod \left\lceil \frac{K}{\alpha} \right\rceil\right) \\
        & = \bigl(t - (s+1) + 1\bigr) / \left\lceil \frac{K}{\alpha} \right\rceil
        \geq \left\lfloor \frac{\alpha}{K} (t - s) \right\rfloor.
    \end{align*}
    Hence we have the result of Proposition~\ref{prop:6:EnoughSamples}.
\end{proof}

\paragraph{Other forced exploration schemes.}
%
We present in Appendix~\ref{sec:6:exploringDifferentForcedExplorationMechanisms} numerical simulations that compare three different options of forced exploration schemes.
We compare the uniformly random exploration used by \CUSUMUCB, against the deterministic scheme used in Algorithm~\ref{algo:6:GLRklUCB}, and against a more complicated scheme based on \emph{tracking}.
The sum-up of these simulations is that it is enough to use the deterministic scheme to have an efficient change-point detection algorithm, and an efficient \GLRklUCB{} policy.
As it is the simplest and less numerically costly one, this is the one we chose for Algorithm~\ref{algo:6:GLRklUCB}.

@@

This Section starts by giving a unified analysis of regret of the two variants of our algorithm, then we give the finite-time regret bounds for the two variants, with their proofs, which both follow the skeleton of the unified analysis.
% We highlight that our results are all non-asymptotic, contrarily to what is given for \MUCB{} \cite{CaoZhenKvetonXie18} and \CUSUMUCB{} \cite{LiuLeeShroff17}.

@@

\begin{assumption}[Enough samples between two global change-points]
    \label{ass:6:LongPeriodsGlobal}
    Define
    \begin{equation}\label{def:6:DelayGlobal}
        d^{(k)} = d^{(k)}(\alpha,\delta) = \left\lceil \frac{4K}{\alpha\left(\Delta^{(k)}\right)^2}\beta(T,\delta) + \frac{K}{\alpha} \right\rceil.
    \end{equation}
    %
    Then we assume that for all $k \in \{1,\dots,\Upsilon_T\}$,
    $\tau^{(k)} - \tau^{(k-1)} \geq 2\max (d^{(k)},d^{(k-1)})$.
\end{assumption}

@@

We first introduce some notation for the proof:
let $\widehat{\tau}^{(k)}$ be the $k$-th change detected by the algorithm, leading to the $k$-th (full) restart and let $\widehat{\tau}(t)$ be the last time before $t$ that the algorithm restarted.
We denote $n_i(t) = \sum_{s=\tau(t)+1}^{t}\indic(A_s = i)$ the number of selections of arm $i$ since the last (global) restart, and $\widehat{\mu}_i(t) = \frac{1}{n_i(t)}\sum_{s=\tau(t)+1}^{t}X_{i,s}\indic(A_s = i)$ their empirical average (if $n_i(t) \neq 0$).

@@

Some new notations are needed to state a regret bound for \GLRklUCB{} using local changes. We let $\tau_i^{(\ell)}$ denote the position of the $\ell$-th change point \emph{for arm $i$}: $\tau_i^{(\ell)} = \inf \{ t > \tau_i^{(\ell - 1)} : \mu_i(t) \neq \mu_i(t+1)\}$,
with the convention $\tau_i^{(0)}=0$, and let $\overline{\mu}_i^{(\ell)}$ be the $\ell$-th value for the mean of arm $i$, such that $\forall t \in [\tau_i^{(\ell-1)}+1, \tau_i^{(\ell)}], \ \ \mu_i(t) = \overline{\mu}_i^{(\ell)}$. We also introduce the gap $\Delta_i^{(\ell)} = \overline{\mu}_i^{\ell} - \overline{\mu}_i^{\ell-1} > 0$.
% ($\ell$ is used instead of $k$ to distinguish notations between local and global changes)

\begin{assumption}[Enough samples between two local change-points]
    \label{ass:6:LongPeriods}
    Define
    \begin{equation}\label{def:6:DelayLocal}
        d_i^{(\ell)} = d_i^{(\ell)}(\alpha,\delta) = \lceil \frac{4K}{\alpha\left(\Delta_i^{(\ell)}\right)^2}\beta(T,\delta) + \frac{K}{\alpha}\rceil.
    \end{equation}
    %
    We assume that for all arm $i$ and all $\ell \in \{1,\dots,\NCi\}$, $\tau_i^{(\ell)} - \tau_i^{(\ell-1)} \geq 2\max (d_i^{(\ell)},d_i^{(\ell-1)})$.
\end{assumption}

@@

% ----------------------------------------------------------------------------
\subsection{Interpretation of the results}
\label{sub:6:interpretationRegretBounds}

>>>>

The Generalized Likelihood Ratio statistic for this test is defined by
\begin{equation}\label{eq:6:firstDefGLRT}
    \GLR(n) :=\frac{\sup\limits_{\mu_0,\mu_1,\tau < n}\ell(X_1, \ldots, X_n ; \mu_0,\mu_1,\tau)}{\sup\limits_{\mu_0}\ell(X_1, \ldots, X_n ; \mu_0)},
\end{equation}
%
where $\ell(X_1, \ldots, X_n ; \mu_0)$ and $\ell(X_1, \ldots, X_n ; \mu_0,\mu_1,\tau)$ respectively denote the likelihoods of the first $n$ observations under a model in $\cH_0$ and $\cH_1$.
High values of this statistic tend to indicate rejection of $\cH_0$ over $\cH_1$.
%
By using the form of the likelihood for Bernoulli distributions, this statistic can be written with the binary relative entropy $\kl$, defined for $x,y\in[0,1]$ (with the usual convention that $t \log(t) = 0$ if $t=0$).
%
\begin{equation}\label{eq:6:BernoulliDivergence}
    \kl(x,y) := x \ln\left(\frac{x}{y}\right) + (1-x)\ln\left(\frac{1-x}{1-y}\right).
\end{equation}
%
Indeed, we show in Appendix~\ref{app:6:GLR_with_kl} that in any one-dimensional exponential family, we have
\begin{equation}\label{eq:6:secondDefGLRT}
    \GLR(n) = \exp\Bigl[ \sup_{s \in [2,n-1]} \left[s \times \kl\left(\widehat{\mu}_{1:s},\widehat{\mu}_{1:n}\right) + (n-s) \times \kl\left(\widehat{\mu}_{s+1:n},\widehat{\mu}_{1:n}\right)\right], \Bigr].
\end{equation}
%
where for $k \leq k'$, $\widehat{\mu}_{k:k'} := \frac{1}{k - k' + 1} \sum_{l=k}^{k'} X_l$ denotes the average of the observations $X_l$ collected between the instants $k$ and $k'$.
%
This motivates the following definition of the Bernoulli GLRT change point detector.
Note that the $\exp$ appearing in the \eqref{eq:6:secondDefGLRT} is absorbed as a $\log$ in the threshold function $\beta$.

@@

\begin{lemma}\label{lem:6:Delay}
    Let be $\bP_{\mu_0,\mu_1,\tau}$ a probabilistic model under which $X_t \in [0,1]$, and $X_t$ has an average of $\mu_0$ for all $t \leq \tau$, and $\mu_1$ for all $t > \tau$, with $\mu_0 \neq \mu_1$,
    and let $\Delta = |\mu_0 - \mu_1|$.
    Then the B-GLRT test given in Definition~\ref{def:6:GLRDef} satisfies
    \begin{equation}
        \bP_{\mu_0,\mu_1,\tau} (\widehat{\tau}_{\delta} \geq \tau + u) \leq \exp\left( -\frac{2\tau u}{\tau + u}\left(\max\left[ 0, \Delta - \sqrt{\frac{\tau + u}{2\tau u} \beta(\tau + u,\delta)} \right]\right)^2 \right).
    \end{equation}
\end{lemma}
%
\begin{proof}
    It is actually not used as such in the proofs of the regret upper bound given below, but a similar result is used and proved in the proofs.
\end{proof}

@@

Also note that, as any test based on scan-statistics, the B-GLRT can be costly to implement as at every time step, it considers all previous time steps as a possible positions for a change-point. Thus, in practice the following adaptation may be interesting, based on down-sampling the possible time steps:
%
\begin{equation}\label{def:6:GLRTricks}
    \widetilde{\tau}_{\delta} = \inf \left\{ n \in \cN : \sup_{s \in \cS_n} \left[s \times \kl\left(\widehat{\mu}_{1:s},\widehat{\mu}_{1:n}\right) + (n-s) \times \kl\left(\widehat{\mu}_{s+1:n},\widehat{\mu}_{1:n}\right)\right] \geq \beta(n,\delta)\right\},
\end{equation}
%
for subsets $\cN$ and $\cS_n$. Following the proof of Lemma~\ref{lem:6:FalseAlarm}, we can easily see that this variant enjoys the same false-alarm control.
However, the detection delay may be slightly increased.
In Appendix~\ref{sub:6:IdeasOptimizations} we show that using these practical speedups has little impact on the regret.

@@

% %-----------------------------------------------------------------------------
% \paragraph{Details about the algorithm.}

@@

\begin{proof}
    We consider one arm $i\in\{1,\dots,K\}$, and when the \GLRklUCB{} algorithm is running,
    we consider two time steps $s\leq t\in\N^*$, chosen between two restart times for that arm $i$.
    Lines~$3$-$4$ state that $A_u = u \mod \lceil \frac{K}{\alpha} \rceil$
    if $u \mod \lceil \frac{K}{\alpha} \rceil \in \{1,\dots,K\}$,
    % (see Algorithm~\ref{algo:6:GLRklUCB} for details).
    thus we directly find
    \begin{align*}
        n_i(t) - n_i(s)
        & = \sum_{u=s+1}^{t} \indic(A_u = i) \\
        & \geq \sum_{u=s+1}^{t} \indic\left(A_u = i, A_u = u \mod \left\lceil \frac{K}{\alpha} \right\rceil\right) \\
        & \geq \sum_{u=s+1}^{t} \indic\left(i = u \mod \left\lceil \frac{K}{\alpha} \right\rceil\right) \\
        & = \bigl(t - (s+1) + 1\bigr) / \left\lceil \frac{K}{\alpha} \right\rceil
        \geq \left\lfloor \frac{\alpha}{K} (t - s) \right\rfloor.
    \end{align*}
    % Hence we have the result of Proposition~\ref{prop:6:EnoughSamples}.
\end{proof}

\paragraph{Other forced exploration schemes?}
%
We present in Appendix~\ref{sec:6:exploringDifferentForcedExplorationMechanisms} numerical simulations that compare three different options of forced exploration schemes.
We compare the uniformly random exploration used by \CUSUMUCB, against the deterministic scheme used in Algorithm~\ref{algo:6:GLRklUCB}, and against a more complicated scheme based on \emph{tracking}.
The sum-up of these simulations is that it is enough to use the deterministic scheme to have an efficient change-point detection algorithm, and an efficient \GLRklUCB{} policy.
As it is the simplest and less numerically costly one, this is the one we chose for Algorithm~\ref{algo:6:GLRklUCB}.

@@

This Section starts by giving a unified analysis of regret of the two variants of our algorithm, then we give the finite-time regret bounds for the two variants, with the proof of the first one (the second one is proved in Appendix~\ref{proof:6:mainRegretBound}), which both follows the skeleton of the unified analysis.

@@

\begin{assumption}[Enough samples between two global change-points]
    \label{ass:6:LongPeriodsGlobal}
    Define
    \begin{equation}\label{def:6:DelayGlobal}
        d^{(k)} = d^{(k)}(\alpha,\delta) = \left\lceil \frac{4K}{\alpha\left(\Delta^{(k)}\right)^2}\beta(T,\delta) + \frac{K}{\alpha} \right\rceil.
    \end{equation}
    %
    Then \emph{we assume} that for all change $k \in \{1,\dots,\Upsilon_T\}$,
    $\tau^{(k)} - \tau^{(k-1)} \geq 2\max (d^{(k)},d^{(k-1)})$.
\end{assumption}

@@

We first introduce some notations for the proof:
let $\widehat{\tau}^{(k)}$ be the $k$-th change detected by the algorithm, leading to the $k$-th (full) restart and let $\widehat{\tau}(t)$ be the last time before $t$ that the algorithm restarted.
We denote $n_i(t) = \sum_{s=\tau(t)+1}^{t}\indic(A_s = i)$ the number of selections of arm $i$ since the last (global) restart, and $\widehat{\mu}_i(t) = \frac{1}{n_i(t)}\sum_{s=\tau(t)+1}^{t}X_{i,s}\indic(A_s = i)$ their empirical average (if $n_i(t) \neq 0$).

@@

Some new notations are needed to state a regret bound for \GLRklUCB{} using local changes,
and to distinguish notations between local and global changes, we denote $\ell$ instead of $k$ for the indexes of change-points.
We let $\tau_i^{(\ell)}$ denote the position of the $\ell$-th change point \emph{for arm $i$}: $\tau_i^{(\ell)} = \inf \{ t > \tau_i^{(\ell - 1)} : \mu_i(t) \neq \mu_i(t+1)\}$,
with the convention $\tau_i^{(0)}=0$, and let $\overline{\mu}_i^{(\ell)}$ be the $\ell$-th value for the mean of arm $i$, such that $\forall t \in [\tau_i^{(\ell-1)}+1, \tau_i^{(\ell)}], \ \ \mu_i(t) = \overline{\mu}_i^{(\ell)}$.
We also introduce the gap $\Delta_i^{(\ell)} = \overline{\mu}_i^{\ell} - \overline{\mu}_i^{\ell-1} > 0$.

\begin{assumption}[Enough samples between two local change-points]
    \label{ass:6:LongPeriods}
    Define
    \begin{equation}\label{def:6:DelayLocal}
        d_i^{(\ell)} = d_i^{(\ell)}(\alpha,\delta) = \lceil \frac{4K}{\alpha\left(\Delta_i^{(\ell)}\right)^2}\beta(T,\delta) + \frac{K}{\alpha}\rceil.
    \end{equation}
    %
    Then \emph{we assume} that for all arm $i$ and all change-point of that arm $\ell \in \{1,\dots,\NCi\}$, $\tau_i^{(\ell)} - \tau_i^{(\ell-1)} \geq 2\max (d_i^{(\ell)},d_i^{(\ell-1)})$.
\end{assumption}

@@

% ----------------------------------------------------------------------------
\subsection{Interpretation and comparison of the results}
\label{sub:6:interpretationRegretBounds}
